---
title: "ROC Curves"
subtitle: ""
author: "FCA Collin, Ph.D."
date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
tags: ["ROC", "graph", "ggplot2"] 
weight: -210110
---

_2021-01-11, FCA Collin (update 2021-01-12)_

---

<!--more-->
```{r setup_knitr, echo = FALSE, message = FALSE, warning = FALSE}
fig.height <- 5;
fig.width <- 16/9 * fig.height;
knitr::opts_chunk$set(
  fig.width  = fig.width,
  fig.height = fig.height,
  dev = c('png', 'pdf', 'svg'),
  fig.path = "img/"
)
```

# Presentation

The Receiver Operating Characteristic (ROC) is a general representation of a
binary classifier; it accounts for:

- the sensitivity (the proportion of real positive case detected),
- the false positive rate (1 - sensitivity),
- the general performance via the area under the ROC curve (AUC).

```{r, echo = FALSE, results = "hide"}
load(file = file.path("data", "ROC.RData"))
```

The classifier for the example identifies tumour tissues (yes/no) in the
case of lung cancer (LC) or eventually identifies the
Squamous Cell Carcinoma (SCC) histological subtype.
The dataset used for the graphic requires at least columns for the sensitivity,
the false-positive rate and a decision rule represented by a threshold
proportion, in that case varying from 0 to 1. Additionally, information about
the classifier itself can include the model identification (LC, SCC), as
well as the AUC (+ confidence interval estimation) as a overall evaluation
of the models.

# Data

Lets `dtaplot` being as example dataset such as:

```{r, result = "asis", echo = FALSE}
dtaplot <- subset(dtaplot, Diag %in% c("1) LC", "2) SCC"));
knitr::kable(
  head(dtaplot),
  caption = "First row of the dataset behind the ROC curves.",
  row.names = FALSE
)
```

Some decision rules (threshold) were of interest:

- `Min. error` is the threshold for which the classifier overall error
  was at its minimum.
- `Sens. 90%`: has its better not to miss true positive patients,
  the threshold may be determined so as to catch 90% of true positive
  patient (sensibility), eventually at a cost in terms of overall error as
  it then automatically increase the false-alarm rate.

Lets `threshold` being the supplementary data characterising this two decision
rules:

```{r, result = "asis", echo = FALSE}
threshold <- rbind(
  within(threshold50, Rational <- "Min. error"),
  within(threshold, Rational <- "Sens. 90%")
)
threshold <- subset(threshold, Diag %in% c("1) LC", "2) SCC"));
knitr::kable(
  threshold,
  caption = "`Threshold, the decision rules.",
  row.names = FALSE
)
```

# Graphic

The example is based on the package `ggplot2`, plus the optional `ggthemr`
which provides graphical themes, for instance the theme`flat`.

```{r, message = FALSE}
library(ggplot2)
if(require(ggthemr)) ggthemr::ggthemr("flat")
```

## Basic

The minimal ROC representation is simply a line plot representing the
sensitivity as a function of the false alarm rate, for one or the
other model.

```{r figure_01}
{
  ggplot(
    data = dtaplot,
    mapping = aes(
      x = FalseAlarm,
      y = Sensitivity,
      color = Diag
    )
    ) + geom_line(
    lwd = 1
    ) + facet_grid(
    . ~ Diag
  )
}
```

## Aesthetic improvement

Aesthetic can help improving the reading:

- the area under the curve is of interest, therefore it can be filled.
- the bisector delimits a model performing as good as a decision made
  flipping a coin (the reference model).
- sensibility and false-alarm rate are define from 0 to 1, the length of
  this two axis should equal.

```{r}
{
  ggplot(
    data = dtaplot,
    mapping = aes(
      x = FalseAlarm,
      y = Sensitivity,
      ymax = Sensitivity,
      fill = Diag,
      color = Diag
    )
    ) + geom_ribbon(
    ymin = 0, alpha = .5, color = NA
    ) + geom_line(
    lwd = 1
    ) + geom_abline(
    slope = 1, intercept = 0, col = "gray50", lwd = 2, lty = 2
    ) + xlab(
    "False Alarm Rate (1 - specificity)"
    ) + facet_grid(
    . ~ Diag
    ) + theme(
    asp = 1
  )
}
```

## Other data for more annotations

The use of the model for diagnostic demands to define a threshold, various
rational can be used, in the example two thresholds were defined:
the minimal error, the 90% detection of positive case. Points can identify
this threshold and performance on the ROC curve.

```{r}
{
  ggplot(
    data = dtaplot,
    mapping = aes(
      x = FalseAlarm,
      y = Sensitivity,
      ymax = Sensitivity,
      fill = Diag,
      color = Diag
    )
    ) + geom_ribbon(
    ymin = 0, alpha = .5, color = NA
    ) + geom_line(
    lwd = 1
    ) + geom_abline(
    slope = 1, intercept = 0, col = "gray50", lwd = 2, lty = 2
    ) + xlab(
    "False Alarm Rate (1 - specificity)"
    ) + facet_grid(
    . ~ Diag
    ) + theme(
    asp = 1
    ) + geom_point(
    data = threshold,
    mapping = aes(shape = Rational, colour = Rational),
    size = 3
  )
}
```

The plot can be further personalised manipulating the theme locally to
address for instance the positioning of the legend and other settings.

```{r}
{
  ggplot(
    data = dtaplot,
    mapping = aes(
      x = FalseAlarm,
      y = Sensitivity,
      ymax = Sensitivity,
      fill = Diag,
      color = Diag
    )
    ) + geom_ribbon(
    ymin = 0, alpha = .5, color = NA
    ) + geom_line(
    lwd = 1
    ) + geom_abline(
    slope = 1, intercept = 0, col = "gray50", lwd = 2, lty = 2
    ) + xlab(
    "False Alarm Rate (1 - specificity)"
    ) + facet_grid(
    . ~ Diag
    ) + geom_point(
    data = threshold,
    mapping = aes(shape = Rational, colour = Rational),
    size = 3
    ) + theme(
    asp = 1, legend.position = "bottom",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 8),
    legend.background = element_rect(fill = "transparent"),
    plot.background = element_rect(fill = alpha("white", .5), colour = 'white')
  )
}
```

## Final

Finally, as a last piece of annotation, the AUC given with its confidence
interval may help for further comparison.

```{r}
{
  ggplot(
    data = dtaplot,
    mapping = aes(
      x = FalseAlarm,
      y = Sensitivity,
      ymax = Sensitivity,
      fill = Diag,
      color = Diag
    )
    ) + geom_ribbon(
    ymin = 0, alpha = .5, color = NA
    ) + geom_line(
    lwd = 1
    ) + geom_abline(
    slope = 1, intercept = 0, col = "gray50", lwd = 2, lty = 2
    ) + xlab(
    "False Alarm Rate (1 - specificity)"
    ) + facet_grid(
    . ~ Diag
    ) + geom_point(
    data = threshold,
    mapping = aes(shape = Rational, colour = Rational),
    size = 3
    ) + theme(
    asp = 1, legend.position = "bottom",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 8),
    legend.background = element_rect(fill = "transparent"),
    plot.background = element_rect(fill = alpha("white", .5), colour = 'white')
    ) + geom_label(
    data = aggregate(auc.ci ~ Diag + ntree, data = dtaplot, unique),
    mapping = aes(label = auc.ci, ymax = NULL, fill = NULL, color = NULL ,
      x = .5, y = .5
    )
  )
}
```

```{r}
sessionInfo()
```

[modeline]: # ( vim: set foldlevel=0 spell spelllang=en_gb: ) 
