---
title: "Beat The Code"
subtitle: ""
author: "FCA Collin, Ph.D."
date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
weight: -210311
---

_2021-03-11, FCA Collin_

---

<!--more-->
```{r setup_knitr, echo = FALSE, message = FALSE, warning = FALSE}
fig.height <- 5;
fig.width <- 16/9 * fig.height;
knitr::opts_chunk$set(
  fig.width  = fig.width,
  fig.height = fig.height,
  dev = c('png', 'pdf', 'svg'),
  fig.path = "img/"
)
```


# `Tidyverse`, Not a Golden Hammer

The Law of the instrument describes a cognitive bias:

> "I call it the law of the instrument, and it may be formulated as follows:
> Give a small boy a hammer, and he will find that everything he encounters
> needs pounding." (Abraham Kaplan, 1964)

It was identified as an _AntiPattern_, a programming practice to be avoided
(William Brown et al, 1998). One of the pit fall is expressed as:

> "the tendency of jobs to be adapted to tools, rather than adapting tools to
> jobs" (Silvan Tomkins, 1963).


The `tidyverse` package helps end-user in R-coding delimited
statistic tasks. It is a very good idea to use it if your purpose is to
walk through an analysis from a point A (the dataset) to the point B 
(the result) for procedures of limited complexity. Indeed, with a limited
number of human-readable functions you can get the expected result while helping
the future reader to follow the procedure.
However, as soon as you want to resolve statistical problems in a more systemic
way, by creating functions that will help you to get your result in a more
concise (because accurate) code and tested for and documented and robust,
`tidyverse` is not the most suitable choice.

The package vignette itself enclose a clear disclaimer about the package
rational:

> *"the biggest difference is in priorities: base R is highly focussed on
> stability, whereas the tidyverse will make breaking changes in the search for
> better interfaces."*
> [_Welcome to the Tidyverse_ vignette, 2019](https://cran.r-project.org/web/packages/tidyverse/vignettes/paper.html)

The trade-off between stability and interface evolution is also acknowledged:

> Do you expect the tidyverse to be the part of core R packages some day?
>
> **Hadley Wickham:** *"It’s extremely unlikely because the core packages are
> extremely conservative so that base R code is stable, and backward compatible.
> I prefer to have a more utopian approach where I can be quite aggressive
> about making backward incompatible changes while trying to figure out a
> better API."*
> [quora](https://www.quora.com/q/quorasessionwithhadleywickham/Do-you-expect-the-tidyverse-to-be-the-part-of-core-R-packages-some-day-1)


There is no doubt that `tidyverse` is a set of high quality tools, but it is
designed to serve some purpose: easy and highly readable code at the cost of
stability which is a strategy which can't serve all developments. Besides,
the over reliance of craftsperson on a known tool, brings to see the challenge
not as it is but how it fit to the tool.
As a matter of fact, an over reliance on `tidyverse` risk to introduce a
cognitive bias,
increasing the risk of of deviation from initial target as fitting your purpose
to the problem instead of making the method suitable to answer specific
question. The over-reliance can be evidence by a large `tidyverse` block
which have obviously lost the main sells argument of the package: readability.

In order to increase the range of possible ways to address a question, so
as to minimise the risk of programming cognitive bias, it is good to
demonstrate alternatives to the `tidyverse` approach put in some context,
and present the R base alternative. There will be a trade-off switching from
one to the other about readability, performance and code stability. But,
maybe, this will also help thinking about different approaches to address
statistical problems.




## Filter and Select

```{r, results = "hide", message = FALSE}
library(tidyverse)
```

Tidyerse Non-Standard Evaluation:

```{r}
iris %>%
  filter(Species == "setosa") %>%
  select(Sepal.Width, Sepal.Length) %>%
  head
```

Base Non-Standard Evaluation:

```{r}
sel <- subset(
  iris,
  subset = Species == "setosa",
  select = c(Sepal.Width, Sepal.Length)
)
head(sel)
```

Base `data.frame` accessors:

```{r}
sel <- iris[
  iris$Species == "setosa",
  c("Sepal.Width", "Sepal.Length")
  ]
head(sel)
```


## Mutate


```{r, eval = FALSE}
df[df$age > 90, ] <- NA
```

##

```{r}
sessionInfo()
```

[modeline]: # ( vim: set foldlevel=0 spell spelllang=en_gb: ) 
