---
title: "A Bit of Style"
subtitle: ""
author: "FCA Collin, Ph.D."
date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
weight: -210415
---

_2021-03-11, FCA Collin_

---

<!--more-->
```{r setup_knitr, echo = FALSE, message = FALSE, warning = FALSE}
fig.height <- 5;
fig.width <- 16/9 * fig.height;
knitr::opts_chunk$set(
  fig.width  = fig.width,
  fig.height = fig.height,
  dev = c('png', 'pdf', 'svg'),
  fig.path = "img/"
)
```


# Style Advises

## Use Comment Appropriately


Main reason to comment:

- Code Readability
- Explanation of the code or Metadata of the project
- Temporarily prevent execution of code
- To include resources

<https://style.tidyverse.org/syntax.html> (210415):

In data analysis code, use comments to record important findings and analysis
decisions. If you need comments to explain what your code is doing,
consider rewriting your code to be clearer. If you discover that you have
more comments than code, consider switching to R Markdown.

### How to comment

<https://style.tidyverse.org/functions.html> (210415):

- Explain the _why_ and not the _what_ or _how_
- Comments should be in sentence case, and only end with a full stop if they
  contain at least two sentences

### Down-side of (bad) comment practice.

- Focus and cognitive load: stops screening code because
  _maybe that is important_ (and 99% of case is not).
- Hides what's important: increase the length of the code. If, eventually,
  it is understood by the programmer who commented it, it is noise to
  every one else. It is better to
  keep the code to the necessary piece and as, along with other measures,
  it will improve code maintenance by preventing to have to look
  for the needle in the haystack.
    + Alternative: if all this comment is necessary, use literate
      programming such as rmarkdown.
- The commented code does not evolve and may become out of date: and this
  is a problem as all the code should work.
- Unfinished job: the comment can be an alternative code as the programmer
  thinks "_maybe that could be another valid approach_". The programmer
  is still the better placed to decided what is better, or may ask advises,
  but the job will be finish when finally ruling the decision. Alternative:
    + if it is hard to decide, take the time to weight the decision,
      eventually ask another colleague.
    + if the decision must be documented: consider literate programming with
      rmarkdown.
- Someone else will delete it. As ambiguous, someone may decide to delete the
  commented code. If this was really informative, that will be a loss which
  could have been prevented (i.e. literate programming, unit tests). Some
  programmers think:
     + _I’ll Delete Your Commented Code Without Reading It and I’m Not Sorry_
       <https://blog.submain.com/delete-commented-code-without-reading/>

Around the Web:

- <https://kentcdodds.com/blog/please-dont-commit-commented-out-code>
- <https://agiletribe.wordpress.com/2015/12/26/never-leave-commented-code-in-the-source/>
- <https://blog.submain.com/delete-commented-code-without-reading/>

Few quotes:

- _The cost of that commented code always outweighs the benefit._
- _Deleting commented code is an easy and effective way to improve the code
  base, without any risk of negative consequences._

##

```{r}
sessionInfo()
```

[modeline]: # ( vim: set foldlevel=0 spell spelllang=en_gb: ) 
